Stack Safety as a Security Property

### 2021-04-06 tex/main.tex
a few comments and rewords on 1-2

But what {\em is} this stack safety of which they speak? Precise formal
definitions are mostly lacking: rather, the notion is commonly defined in a
negative way, by enumerating the bad things that a given enforcement mechanism
prevents, such as corruption of return addresses, buffer overflows, or
use of uninitialized variables.
%\bcp{Does use of unitialized variables lead to stack safety attacks?\apt{sure, via buffer overflow}}
%with the notion usually being
%defined negatively: through examples of stack-based exploits
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ...
Indeed, we are only aware of one quite recent attempt to frame a positive definition, by
\citet{Skorstengaard+19}, which defines stack safety as the conjunction of two properties:
{\em local state encapsulation} and {\em well-bracketed control flow}.
These properties are formalized by defining a capability-based machine
that ``satisfies [stack safety] by construction.''
%
While this work represents an important step in the right direction, we believe
there is still room to craft an operational characterization
of stack safety. In other words, we seek formal criteria that can be applied
to a mechanism for protecting stack-based low-level programs to judge
whether or not it actually ``guarantees stack safety.'' We begin with a strong
local state encapsulation property that closely models the behavior of eager
micro-policies from \citet{DBLP:conf/sp/RoesslerD18}. This ``eager'' property
is too strong to characterize stack safety; reasonable enforcement mechanisms
exist that violate it. One such mechanism is a ``lazy'' tag policy, similar to
the one in \citet{DBLP:conf/sp/RoesslerD18}. We weaken local state encapsulation
to support this technique and argue that the new property, defined relative to an
arbitrary notion of observation, is weak enough to characterize stack safety.

While our criteria are designed with PIPE micro-policies in mind, they are
agnostic about many implementation details and should be applicable to other
systems with minor adjustment.
%
Our key contribution is an operational model of local state encapsulation
using the technical framework of language-based security, with eager and lazy variants,
together with a simple characterization of well-bracketed control flow.

@@

How does the programmer rely on stack safety? The stack makes implicit promises about
the behavior of function calls. First, that when a call is finished, the caller can continue
from where it left off with its local state unchanged. And second, that the call is
self-contained, its behavior determined only by a predictable interface -- at minimum
parameters and global variables, but never local state.

@@

Implicit in both expectations are a predictable control flow, termed {\em well-bracketed
control flow} (WBCF) by \citeauthor{Skorstengaard+19}. Calls jump to a consistent entry
point and return to the call site. Then the data portion of the first expectation is for
{\em integrity}: that a caller's local data is protected from modification by its callee.
This is vital for local reasoning about the behavior of the caller. The security implications
are significant: a caller that branches on a local variable could be manipulated by its
callee into unexpected behavior without integrity.

Alongside integrity, we have {\em confidentiality}, in which a caller's data is protected
from leakage by its callee. As a reasoning principle, confidentiality tells us that to
predict the behavior of a call we need only consider a subset of the machine state.
Rearranging local data will have no impact. And in security terms, we can guarantee
the secrecy of local data that we haven't passed even when calling a badly behaved function.
Together, integrity and confidentiality comprise \citeauthor{Skorstengaard+19}'s
{\em local state encapsulation} (LSE) property.

@@

To phrase LSE and WBCF as security properties requires an explicit attacker model:
a clear understanding of who is being protected and from what. Most existing
stack protection mechanisms envisage scenarios where a callee is
influenced by external inputs to corrupt the stack in some way. To
avoid getting into the details of exactly how this happens when
formalizing our properties, we use an even stronger model: when we
state the protections that a function can rely on when it makes a
call, we assume that the callee can execute arbitrary instructions
until it returns. Thus we have no need to differentiate safe code
from adversarial - all functions, and indeed all activations of the same function,
are mutually distrusting.
%
Even more fundamentally, we need to know when a call or a return is taking
place. We assume that calls and returns are annotated in the code, so that
we can track the point at which control transfers from the protected caller
to the attacking callee -- and when that callee in turn becomes a protected
caller in its own right.

@@

First we introduce this model for a single simple stack with no sharing between callers and
callee. Then we expand it to allow components to be passed, granting a callee one time
access to the caller's data. We do not model more sophisticated sharing, because to do
so realistically would require a separate protection scheme, such as a capability
model. While this is quite feasible in a tag-based enforcement mechanism, it is properly
a separate policy outside the scope of stack safety. (The combination of such policies is
a strength of tag-based approaches.)

Finally, we extend the model to a coroutine system featuring finitely-bounded stacks. Though
simple, this extension is a step toward concurrency and demonstrates that the model is flexible
enough to handle more sophisticated styles of control-flow.

After formally defining stack safety, we show that the definition is
enforceable using an existing tag-based enforcement
mechanism~\citep{DBLP:conf/sp/RoesslerD18}. We first consider
\citeauthor{DBLP:conf/sp/RoesslerD18}'s conservative Depth Isolation policy.
This policy testably enforces our {\em eager} properties, in which integrity is respected
at every step of a program, and each call respects integrity at each of its internal steps.
These properties are very strong, and our policies enforce them; they also have the advantage
that they ``fail faster'' and are therefore suitable for efficient testing.

The downside of the Depth Isolation policy is that it is rather expensive to implement.
In the same work, \citeauthor{DBLP:conf/sp/RoesslerD18} propose a different mechanism
that is more efficient by virtue of being lazy in its enforcement.
Instead of enforcing the integrity and confidentiality of the stack at the
exact moment when a callee returns, this policy signals violations when the
caller actually accesses data that the callee has written into its stack
frame. \ifaftersubmission\bcp{We could also hint at the lazy confidentiality
  property in the same way...}\fi
The lazy policy does not enforce our eager integrity property, and to our
knowledge there is no formal statement of the properties that it ought to enforce.
We propose {\em lazy} stack-safety properties, which capture the desired behavior of a
policy that detects integrity violations between when they occur and when they make
a visible impact. Intuitively this can be phrased, ``No visible reads after dangerous
writes.'' Unfortunately \citeauthor{DBLP:conf/sp/RoesslerD18}'s existing
``Lazy Tagging and Clearing'' policy does not quite enforce this property, as we show in
testing, but formalizing the ideal property allows us to see how it falls short.
We provide a variant of ``Lazy Tagging and Clearing'' that does enforce lazy integrity
at some performance cost. We prove (in Coq) that the eager property implies the lazy one.
We argue that any system purporting to enforce stack safety ought to obey lazy stack safety.

@@

In summary, we offer the following contributions:
\begin{itemize}
\item
  % \leo{More? Extensional}\bcp{We might need to invent a word for this---''quasi-extensional''?}
  A novel {\em stack-safety property} capturing the intuitive concepts of strictly enforced
  local state encapsulation and well-bracketed control flow~(\cref{sec:lse-and-wbcf}).
  % These properties rule out common examples of stack memory abuse.
\item
  A property-based randomized testing framework that demonstrates that this
  property can be enforced by the Depth Isolation policy of
  \citet{DBLP:conf/sp/RoesslerD18}, which is based on hardware
  tagging~(\cref{sec:conservative}).
\item
  A {\em lazy} variant of the stack-safety property that can be enforced if the
  Lazy Tagging and Clearing policy of \citet{DBLP:conf/sp/RoesslerD18} is enhanced
  with unique tags on each activation, plus a proof (in Coq) that this
  variant is weaker than the original~(\cref{sec:lazy}).
\item
  Standardized trace-based formulations of both eager and lazy confidentiality and integrity,
  parameterizeable by arbitrary control-flow models and assignments of machine components
  to {\em LSE domains}.
\item
  An LSE-domain model enhanced with passing of function arguments on the stack, with which our
  standardized trace properties extend stack safety with local sharing.
\item
  An LSE-domain model extending the stack to a simple coroutine model, with corresponding LSE
  properties and a {\em well-structured control flow} property that subsumes WBCF for each
  stack as well as guaranteeing legitimate coroutine crossings.
%  This policy does not enforce our strict
%  local state encapsulation property,
%  but a modified version of it using unique per-activation tags enforces
%  an (also provably) weaker lazy variant of the property, which still prohibits
%  the same intuitively stack unsafe behaviors.
%
% \apt{Awkward, but the truth I fear.} \bcp{I agree the phrasing is awkward,
%   but the point we can make is still strong: To make the property easier
%   to enforce, we can slightly weaken its guarantees; the weaker version is
%   a bit more complicated to state but still retains the pragmatic
%%   protections of the original.}\apt{To clarify: Nick's LT\&C policy uses
%   depth-based tags, which do not enforce even our lazy policy. So we need to
%   introduce a version of LT\&C that uses unique activation tags.}

@@

In this very strong attacker model, there is little obvious distinction between
the attacker and the defender. They are separated only by the control transfer
inherent in their calls and returns. Our stack-safety concepts therefore depend
critically on the notions of ``call'' and ``return,'' which are not fully explicit
in machine code (the same instruction opcode may be used in some places as part of
a call or return sequence and in other places for other purposes).
%
We assume we are given a machine language program annotated with the locations of
instructions representing calls, returns, and function entry points. Although this
information would typically come from a compiler together with the machine code,
we do not assume anything about their provenance. We do treat the execution of an
annotated instruction as in some sense intentional on the part of the current
function. No accidental calls or returns, and attackers control their own returns
but cannot change the annotations on the caller's code.

@@

Since we rely so heavily on code annotations, we must ask: what if the annotations
are wrong? What if a call or return occurs that is not annotated, or an annotation
fails to correspond to an actual return? In the former case, the answer is simple --
there has been no call or return, only instruction resembling such an operation, and
therefore no associated change in permissions. This is particularly interesting in the
case of a false return, where control may appear to have returned to the caller, but
the enforcement mechanism must still protect the caller's memory.

The latter case depends on the enforcement mechanism and the nature of the false
annotation. A given enforcement mechanism might place a static well-formedness criterion
on the code it applies to. For instance, our micro-policies expect that call and return
annotations occur on jump instructions. On the other hand, not all jumps are proper
calls or returns, only those preceded by appropriate sequences: in these cases the
bad annotation might appear well formed but the system should either respect the change
in permissions or, if doing so would compromise security, fail-stop rather than
execute the annotated instruction.

@@

@@

Our definitions are largely parametric over the layout of the stack in memory, but for
purposes of explanation we will describe a standard layout. Memory consists of an
instruction region and a stack that grows upwards. (For simplicity,
we do not consider the heap. Integrating a memory-safety property for the
heap in the style of \citet{DBLP:conf/post/AmorimHP18} is a natural extension
that we discuss in~\Cref{sec:relwork}.) The stack has a conventional structure, with
contiguously allocated per-function frames.

@@

\begin{itemize}
\item Examples that appear to violate stack safety, violate at least one of our properties
\item We do not find desireable behaviors that appear stack-safe yet violate our properties
\item Our properties are testably implemented by stack safety micro-policies from the literature,
  and the micro-policies are not significantly more conservative (i.e., policy failstops tend to
  correspond to imminent property violations)
\end{itemize}

>>>>

But what {\em is} the stack safety that these techniques enforce? Precise formal
definitions are mostly lacking: rather, the notion is commonly defined in a
negative way, by enumerating the bad things that a given enforcement mechanism
prevents, such as corruption of return addresses, buffer overflows, or
use of uninitialized variables.
%\bcp{Does use of unitialized variables lead to stack safety attacks?\apt{sure, via buffer overflow}}
%with the notion usually being
%defined negatively: through examples of stack-based exploits
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ...
Indeed, we are only aware of one quite recent attempt to frame a positive definition, by
\citet{Skorstengaard+19}, which defines stack safety as the conjunction of two properties:
{\em local state encapsulation} and {\em well-bracketed control flow}.
These properties are formalized by defining a capability-based machine
that ``satisfies [stack safety] by construction.''
%
While this work represents an important step in the right direction, we believe
there is still room to craft an operational characterization
of stack safety. \apt{Perhaps our rhetoric should be more micro-policy-specific here?}
In other words, we seek formal criteria that can be applied
to a mechanism for protecting stack-based low-level programs to judge
whether or not it actually ``guarantees stack safety.'' We begin with a strong
local state encapsulation property that closely models the behavior of eager
micro-policies from \citet{DBLP:conf/sp/RoesslerD18}. This ``eager'' property
is too strong to characterize stack safety; reasonable enforcement mechanisms
exist that violate it. One such mechanism is a ``lazy'' tag policy, similar to
the one in \citet{DBLP:conf/sp/RoesslerD18}. We weaken local state encapsulation
to support this technique and argue that the new property, defined relative to an
arbitrary notion of observation, is weak enough to characterize stack safety.
\apt{Reads oddly. ``string enough'' is what we want, surely? (While still weak
  enough to be enforced by the lazy policies.)}

While our criteria are designed with PIPE micro-policies in mind, they are
agnostic about many implementation details and should be applicable to other
systems with minor adjustment.
%
Our key contribution is an operational model of local state encapsulation
using the technical framework of language-based security\apt{this idea deserves more prominence}, with eager and lazy variants,
together with a simple characterization of well-bracketed control flow.

@@

How does the programmer rely on stack safety? The stack makes implicit promises about
the behavior of function calls. First, that when a call is finished, the caller can continue
from where it left off with its local state unchanged. And second, that the call is
self-contained, its behavior determined only by a predictable interface -- at minimum
parameters and global variables, but never local state.\apt{This last sentence is not so clear.}

@@

Both expectations imply predictable control flow, termed {\em well-bracketed
control flow} (WBCF) by \citeauthor{Skorstengaard+19}. Calls jump to a consistent \apt{??} entry
point and return to the call site. Then the data portion of the first expectation is for
{\em integrity}: that a caller's local data is protected from modification by its callee.
This is vital for local reasoning about the behavior of the caller. The security implications
are significant: without integrity, a caller that branches on a local variable could be manipulated by its
callee into unexpected behavior.

Alongside integrity, we have {\em confidentiality}, in which a caller's data is protected
from leakage by its callee. As a reasoning principle, confidentiality tells us that to
predict the behavior of a call we need only consider a subset of the machine state.
Rearranging local data will have no impact.\apt{??} And in security terms, we can guarantee
the secrecy of local data that we haven't passed even when calling a badly behaved function.
Together, integrity and confidentiality comprise \citeauthor{Skorstengaard+19}'s
{\em local state encapsulation} (LSE) property.

@@

To phrase LSE and WBCF as security properties requires an explicit attacker model:
a clear understanding of who is being protected and from what. Most existing
stack protection mechanisms envisage scenarios where a callee is
influenced by external inputs to corrupt the stack in some way. To
avoid getting into the details of exactly how this happens when
formalizing our properties, we use an even stronger model: when we
state the protections that a function can rely on when it makes a
call, we assume that the callee can execute arbitrary instructions
until it returns. Thus we have no need to differentiate safe code
from adversarial - all functions, and indeed all activations of the same function,
are mutually distrusting.
%
Even more fundamentally, \apt{``For this model to make sense,'' (??)} we need to know when a call or a return is taking
place. We assume that calls and returns are annotated in the code, so that
we can track the point at which control transfers from the protected caller
to the attacking callee -- and when that callee in turn becomes a protected
caller in its own right.

@@

First we introduce this model for a single simple stack with no sharing between callers and
callee. Then we expand it to allow components\apt{??} to be passed, granting a callee one time
access to the caller's data. We do not model more sophisticated sharing, because to do
so realistically would require a separate protection scheme, such as a capability
model. While this is quite feasible in a tag-based enforcement mechanism, it is properly
a separate policy outside the scope of stack safety. (The combination of such policies is
a strength of tag-based approaches.)

Finally, we extend the model to a coroutine system featuring finitely-bounded stacks. Though
simple\apt{simplistic}, this extension is a step toward concurrency and demonstrates that the model is flexible
enough to handle more sophisticated styles of control-flow.

After formally defining stack safety, we show that the definition is
enforceable using an existing tag-based enforcement
mechanism~\citep{DBLP:conf/sp/RoesslerD18}. We first consider
\citeauthor{DBLP:conf/sp/RoesslerD18}'s conservative Depth Isolation policy.
This policy testably enforces our {\em eager} properties, in which integrity is respected
at every step of a program, and each call respects integrity at each of its internal steps.
These properties are very strong, and our policies enforce them; they also have the advantage
that they ``fail faster'' \apt{than what?} and are therefore suitable for efficient testing.

The downside of the Depth Isolation policy is that it is rather expensive to implement.
In the same work, \citeauthor{DBLP:conf/sp/RoesslerD18} propose a different mechanism
that is more efficient by virtue of being lazy in its enforcement.
Instead of enforcing the integrity and confidentiality of the stack at the
exact moment when a callee returns, this policy signals violations when the
caller actually accesses data that the callee has written into its stack
frame. \ifaftersubmission\bcp{We could also hint at the lazy confidentiality
  property in the same way...}\fi
The lazy policy does not enforce our eager integrity property, and to our
knowledge there is no formal statement of the properties that it ought to enforce.
We propose {\em lazy} stack-safety properties, which capture the desired behavior of a
policy that detects integrity violations between when they occur and when they make
a visible impact. Intuitively this can be phrased as ``no visible reads after dangerous
writes.'' Unfortunately \citeauthor{DBLP:conf/sp/RoesslerD18}'s existing
``Lazy Tagging and Clearing'' policy does not quite enforce this property, as we show in
testing, but formalizing the ideal property allows us to see how it falls short.
We provide a variant of ``Lazy Tagging and Clearing'' that does enforce lazy integrity
at some performance cost. We prove (in Coq) that the eager property implies the lazy one.\apt{do we?
  and is it so interesting anyhow?}
We argue that any system purporting to enforce stack safety ought to obey \apt{``at least'' ?} xslazy stack safety.

@@

In summary, we offer the following contributions:
\begin{itemize}
\item
  % \leo{More? Extensional}\bcp{We might need to invent a word for this---''quasi-extensional''?}
  A novel {\em stack-safety property} capturing the intuitive concepts of strictly enforced
  local state encapsulation and well-bracketed control flow~(\cref{sec:lse-and-wbcf}).
  % These properties rule out common examples of stack memory abuse.
\item
  A property-based randomized testing framework that demonstrates that this
  property can be enforced by the Depth Isolation policy of
  \citet{DBLP:conf/sp/RoesslerD18}, which is based on hardware
  tagging~(\cref{sec:conservative}).
\item
  A {\em lazy} variant of the stack-safety property that can be enforced if the
  Lazy Tagging and Clearing policy of \citet{DBLP:conf/sp/RoesslerD18} is enhanced
  with unique tags on each activation, plus a proof (in Coq) that this
  variant is weaker than the original~(\cref{sec:lazy}).
\item
  Standardized trace-based formulations of both eager and lazy confidentiality and integrity,
  parameterizeable by arbitrary control-flow models and assignments of machine components
  to {\em LSE domains}.
\item
  An LSE-domain model enhanced with passing of function arguments on the stack, with which our
  standardized trace properties extend stack safety with local sharing.
\item
  An LSE-domain model that applies to simple coroutining, with corresponding LSE
  properties and a {\em well-structured control flow} property that subsumes WBCF for each
  stack as well as guaranteeing legitimate coroutine control transfers.
%  This policy does not enforce our strict
%  local state encapsulation property,
%  but a modified version of it using unique per-activation tags enforces
%  an (also provably) weaker lazy variant of the property, which still prohibits
%  the same intuitively stack unsafe behaviors.
%
% \apt{Awkward, but the truth I fear.} \bcp{I agree the phrasing is awkward,
%   but the point we can make is still strong: To make the property easier
%   to enforce, we can slightly weaken its guarantees; the weaker version is
%   a bit more complicated to state but still retains the pragmatic
%%   protections of the original.}\apt{To clarify: Nick's LT\&C policy uses
%   depth-based tags, which do not enforce even our lazy policy. So we need to
%   introduce a version of LT\&C that uses unique activation tags.}

@@

In this very strong attacker model, there is little obvious distinction between
the attacker and the defender. They are separated only by the control transfer
inherent in their calls and returns. Our stack-safety concepts therefore depend
critically on the notions of ``call'' and ``return,'' which are not fully explicit
in machine code (the same instruction opcode may be used in some places as part of
a call or return sequence and in other places for other purposes).
%
We assume we are given a machine language program annotated with the locations of
instructions representing calls, returns, and function entry points. Although this
information would typically come from a compiler together with the machine code,
we do not assume anything about their provenance. We do treat the execution of an
annotated instruction as in some sense intentional on the part of the current
function. No accidental calls or returns, and attackers control their own returns
but cannot change the annotations on the caller's code.\apt{cryptic}

@@

Since we rely so heavily on code annotations, we must ask: what if the annotations
are wrong? First, suppose a call or return occurs that is not annotated.
In this case, the answer is simple:
there has been no call or return\apt{from whose perspective?}, only an instruction resembling such an operation, and
therefore no associated change in permissions. This is particularly interesting in the
case of a false return, where control may appear to have returned to the caller, but
the enforcement mechanism must still protect the caller's memory.

Alternatively, suppose an annotated instruction is not really an actual \apt{``call or'' ?} return.
What happens in this case depends on the enforcement mechanism and the nature of the false
annotation. A given enforcement mechanism might place a static well-formedness criterion
on the code it applies to. For instance, our\apt{or \citeauthor{DBLP:conf/sp/RoesslerD18}'s ? }
micro-policies expect that call and return
annotations occur on jump instructions. On the other hand, not all jumps are proper
calls or returns, only those preceded by appropriate sequences: in these cases the
bad annotation might appear well formed but the system should either respect the change
in permissions or, if doing so would compromise security, fail-stop rather than
execute the annotated instruction.\apt{hard to follow}

@@

\apt{I'm unclear about how many of these assumptions are relevant to the properties
  as opposed to the (micro)policies.}

@@

Our definitions are largely parametric over the layout of the stack in memory, but for
purposes of explanation we will describe a standard layout. Memory consists of an
instruction region and a stack that grows upwards.\apt{if we want to be ``standard''
  we really should have the stack grown downwards.}
(For simplicity,
we do not consider the heap. Integrating a memory-safety property for the
heap in the style of \citet{DBLP:conf/post/AmorimHP18} is a natural extension
that we discuss in~\Cref{sec:relwork}.) The stack has a conventional structure, with
contiguously allocated per-function frames.

@@

\begin{itemize}
\item Examples that appear to violate stack safety, violate at least one of our properties
\item We do not find desireable behaviors that appear stack-safe yet violate our properties
\item Our properties are testably implemented by stack safety micro-policies from the literature,
  and the micro-policies are not significantly more conservative (i.e., policy failstops usually 
  correspond to imminent property violations)
\end{itemize}
